How predictable is extinction? Forecasting species survival at million-year timescales

### 2019-03-07 doc/paper.tex
pandoc-d some alternate formats for paper
seth comments entered

\begin{abstract}
  One of the great promises of paleobiology is that by studying the past we can better predict the future. This promise is particularly pertinent given as risk assessments for some modern species could potentially be improved by examining past extinction patterns and by using paleontological records to establish geographic range and abundance trajectories on geological timescales. Any effort to assess future risk based on past extinctions and range trajectories must address two key questions:  (1) At a given timescale, are geographic range and extinction risk trajectories deterministic (past trends are likely to continue into the future) or Markovian (the future depends only on the present state)? (2) Given knowledge of past extinction/survival patterns and the present geographic ranges of extant taxa, how accurate are extinction risk predictions?  
  To address these questions we analyze the fossil record of Cenozoic planktonic microfossil taxa (foramanifera, radiolarians, diatoms, and calcareous nanoplankton). 
  We analyzed how survival probability changes over time as a function of species age, time of observation, current geographic range, most recent change in geographic range, global temperature average, and the lag of global temperature. 
  %Our best supported model includes the historical covariates, change in geographic range and lag of global temperature, which indicates that the past improves our estimates of the present and future. 
  Our results show that our best performing model has an approximately 78\% median probability of correctly ranking the relative extinction risk any two randomly selected species. 
  We have confidence that our conclusions about our ability to predict species extinction risk in the future with similar accuracy to when we predict species extinction risk in the past because our in-sample model performance measures are approximately equal to out-of-sample performance. 
  Including the historical covariates and allowing their effects to vary over time yields marginally better predictions than not including them. 
  However, the improvement in predictive power by including these historical covariates is modest at best, absent at worst, and ultimately reflects the extremely stochastic nature of species extinction. 
  %Finally, we find support for species' extinction risk increases with age, though the strength of this effect varies among taxonomic groups. 
  %This effect is most pronounced in forams and radiolarians, and less pronounced in diatoms and calcareous nanoplankton. 
  %The greatest source of variance in survival probability is the timing of observation. 
  %Importantly, this result means that the time of an observation is a greater source of variation in survival probability when compared to species age. 
  %We find that including information on a species' change in geographic range size on average improves our predictions of species survival at million-year timescales. 
  %The effect of change in geographic range is much smaller than the effect of current geographic range, and highly variable through time as the effect changes sign and there are times where there is little evidence for any effect of past geographic range. 
  These results imply that at million-year timescales geographic range trajectories are nearly Markovian, perhaps because the processes driving geographic range changes vary on substantially shorter timescales. 
  %The effect of change in geographic range on survival most likely stands for many interacting and unobserved processes which in-turn produce that species' geographic range and its affect on survival.  
  These results reflect the difficulty of estimating species extinction, and that while including historical covariates does improve model performance, that gain is very small. 
  The results of this study reinforce the importance of the promise of paleontology and using the past to predict the future.
\end{abstract}

@@

Being able to predict which species are more likely to go extinct than others is critical for making good conservation decisions to limit the impact of the current biodiversity crisis. We cannot know, however, we do not yet know which species are going to go extinct because this has not happened yet -- it is unobservable. We approach this this problem by analyzing the past in order to predict the future. The fossil record preserves past extinction events, allowing us to develop a predictive model of species extinction based on this record and the properties of the observed species, both extinct and extant \citep{Harnik2012,Finnegan2015}. By assessing the predictive performance of this model on unobserved data, we can quantify how precise our best estimates will be for future extinctions -- we ask the probability that, given two random species, we correctly rank their relative risks of extinction.

@@

For this kind of exercise, we chose to analyze what is the longest continuous and best resolved fossil record -- that of skeletonized marine plantonic microorganisms from the Cenozoic such as Foraminifera, Radiolarians, Diatoms, and calcareous nannofossils (e.g. coccolithophores). This data is available through the Neptune database, an online repository of species occurrences obtained through the Deep Sea Drilling Program and the Ocean Drilling Project \citep{Lazarus1994,SpencerCervato1999}. This database provides abundant samples in space and time, a high degree of temporal resolution for the entirety of the Cenozoic, and has an internally consistent taxonomic identification strategy -- as close to ideal data for this analysis as possible.

Rarely are we able to analyze long periods of geological time at fine resolutions -- below the 5-10 My scale. Due to substantial effort and the unique biology of the system, the microfossil provides us the unique opportunity to analyze ecological and evolutionary patterns at approximately million-year time scales. Typical ``exceptional'' fossil records tend to be of individual taxonomic groups and for rarely longer than 10 million years. The Neptune database records multiple phyla-scale taxonomic groups for over 60 million years, with incredible temporal resolution supported by the various age-models of the deep-sea cores the occurrences are recorded from -- there is no equivalent fossil record. By analyzing patterns of extinction and global occurrence at fine temporal scales, we can better elucidate how well we can predict species extinction at human-relevant scales.

@@

We analyzed microfossil occurrence information which was downloaded from the Neptune Database \url{http://www.nsb-mfn-berlin.de/nannotax} \citep{Lazarus1994,SpencerCervato1999}. All occurrence information was downloaded for calcareous nannofossils, diatoms, foraminifera, and radiolarians -- these occurrences span the entire globe between 120 and 0 million years ago (Mya). This dataset of occurrences was then filtered to just those species which have their first occurrence at most 63 Mya. This choice means that our analysis avoids those taxa which survived the K/Pg boundary, those taxa which arose just after the K/Pg boundary, and means that our occurrence histories line up with the temperature time-series which was used as a predictor of extinction (discussed below).

@@

A taxon's geographic range was calculated for each of the 1 My bins in which it occurred. Geographic range was calculated as the maximum great circle distance on an ellipsoid (i.e. the Earth) between any two occurrences of that species; this measure is also called a geodesic. This distance was measured in kilometers. Geographic range was then log-plus-one transformed, and standardized by zero-centering the data and then dividing by the standard deviation of the distribution of geographic ranges. This standardization means that the analyzed values of geographic range had mean 0 and standard deviation 1. This standardization means that a regression coefficient associated with this covariate describes the change in extinction probability per change in standard deviation of geographic range, that coefficients associated with similarly standardized covariates will be directly comparable in magnitude, and that the interecpt term corresponds to the expected value of the outcome at when geographic range is its average value \citep{ARM}. Change in geographic range between observations was measured from the transformed and standardized geographic range values and not standardized seperately.

Temperature data used as covariates in this analysis are based on Magnesium/Calcium isotope ratios sourced from \citet{Cramer2011}. These elemental ratios are considered more accurate estimates of past global temperature when compared to the frequently used Oxygen isotope-based estimates; this is because Mg/Ca based estimates are not effected by ice-volume and fresh-water input (e.g. meteoric water) which can alter Oxygen isotope ratios without reflecting changes to the climate itself. This property is of particular importance for this analysis as polar ice-caps develop midway through the Cenozoic. Our data source, \citet{Cramer2011}, provides temperature estimates for every 0.1 My from 0 to 63 Mya. We binned these estimates into 1 My intervals as we did with the fossil occurrences. The temperature estimate for each 1 My interval was calculated as the mean of all estimates within that interval. Temperature was then transformed and standardized the in the same manner as geographic range (above). Difference in temperature between between observations was measured from the transformed and standardized temperature values and not standardized seperately.

@@

The in-sample model comparisons are for determining their relative adequacy, or a model's ability to represent the data it was fit to. Comparison between the posterior predictive estimates of in-sample AUC for each of the four models demonstrates that, overall, all of the models have approximately equal in-sample performance (Fig. \ref{fig:auc_hist}). The parameter rich model VP has the greatest median in-sample AUC when compared to the other three models, but there is substantial overlap in their posterior distributions. Additionally, while our parameter rich model VP is possibly the most adequately performing model, the difference or improvement to performance is minimal at best -- all four models have approximately equal in-sample AUC posterior distributions. All of the in-sample AUC estimates from our models are concentrated around an AUC of 0.77 which is interpreted as ``fine but not good'' performance. It is then hard to conclude that there is one ``best'' model which we can rely upon.

@@

In survival analysis, the hazard function describes the instantaneous rate of extinction of a species given its age and relevant covariates. The hazard function is defined as the conditional probability of a species going extinct by the end of the \(t\)-th interval given that it survived up until \(t\) and the relevant covariate information \(X\) for all \(k\) 1 My intervals \citep{Tutz2016}. For the discrete time intervals \(T = 1, \cdots, k\), extinction is defined as \(T = t\). The discrete time hazard function is defined as
\begin{equation}
  \lambda(t | X) = P(T = t | T \geq t, X), \quad t = 1, \cdots, k. \\
  \label{eq:hazard}
\end{equation}

>>>>

\begin{abstract}
  One of the central tenants of conservation paleobiology is that by studying the past we can better predict the future. It has been suggested that risk assessments for extant species could potentially be improved by examining past extinction patterns and by using paleontological records to establish geographic range and abundance trajectories on geological timescales. Such suggestions prompt two key questions: (1) At a given timescale, to what extent are geographic range and extinction risk trajectories deterministic (past trends are likely to continue into the future) versus Markovian (the future depends only on the present state)? (2) At a given point in time, how accurate do extinction risk predictions based on instantaneous geographic ranges and knowledge of past extinction/survival patterns  turn out to be? To address these questions we analyze the exceptionally well-sampled fossil record of Cenozoic planktonic microfossil taxa (foramanifera, radiolarians, diatoms, and calcareous nanoplankton). We examine how survival probability changes over time as a function of species age, time of observation, current geographic range, most recent change in geographic range, climate state, and the change in climate state. Our best performing model has an approximately 78\% median probability of correctly ranking the relative extinction risk any two randomly selected species. In-sample model performance measures are approximately equal to out-of-sample performance measures (assessed with 5-fold cross-validation), suggesting that the determinants of relative extinction risk are fairly consistent throughout the Cenozoic. Although the best-performing model includes time-varying effects and historical covariates (change in range and change in climate state) the improvement in predictive power over models with constant effects and no historical covariates is very minor. Our results imply that at million-year timescales geographic range trajectories are nearly Markovian, perhaps because the processes driving geographic range changes vary on substantially shorter timescales. However, models based on present geographic ranges and past range-risk relationships are likely to be modernly successful at predicting future extinction/survival outcomes. AN important caveat is that human impacts may substantially disrupt range-risk dynamics so that the future will be less predictable than it has been in the past.

  %Our best supported model includes the historical covariates, change in geographic range and lag of global temperature, which indicates that the past improves our estimates of the present and future. 
  %We have confidence that our conclusions about our ability to predict species extinction risk in the future with similar accuracy to when we predict species extinction risk in the past because our in-sample model performance measures are approximately equal to out-of-sample performance. 
  %Including the historical covariates and allowing their effects to vary over time yields marginally better predictions than not including them. 
  %However, the improvement in predictive power by including these historical covariates is modest at best, absent at worst, and ultimately reflects the extremely stochastic nature of species extinction. 
  %Finally, we find support for species' extinction risk increases with age, though the strength of this effect varies among taxonomic groups. 
  %This effect is most pronounced in forams and radiolarians, and less pronounced in diatoms and calcareous nanoplankton. 
  %The greatest source of variance in survival probability is the timing of observation. 
  %Importantly, this result means that the time of an observation is a greater source of variation in survival probability when compared to species age. 
  %We find that including information on a species' change in geographic range size on average improves our predictions of species survival at million-year timescales. 
  %The effect of change in geographic range is much smaller than the effect of current geographic range, and highly variable through time as the effect changes sign and there are times where there is little evidence for any effect of past geographic range. 
  %These results imply that at million-year timescales geographic range trajectories are nearly Markovian, perhaps because the processes driving geographic range changes vary on substantially shorter timescales. 
  %The effect of change in geographic range on survival most likely stands for many interacting and unobserved processes which in-turn produce that species' geographic range and its affect on survival.  
  %These results reflect the difficulty of estimating species extinction, and that while including historical covariates does improve model performance, that gain is very small. 
  %The results of this study reinforce the importance of the promise of paleontology and using the past to predict the future.
\end{abstract}

@@

Being able to predict which species are more likely than others to go extinct over a given time frame is crucial for planning decisions policies to ameliorate the current biodiversity crisis. Some species have been the focus of intense study to define short-term (e.g. decadal to centennial scale) abundance and geographic range trajectories.

We cannot know, however, we do not yet know which species are going to go extinct because this has not happened yet -- it is unobservable. We approach this this problem by analyzing the past in order to predict the future. The fossil record preserves past extinction events, allowing us to develop a predictive model of species extinction based on this record and the properties of the observed species, both extinct and extant \citep{Harnik2012,Finnegan2015}. By assessing the predictive performance of this model on unobserved data, we can quantify how precise our best estimates will be for future extinctions -- we ask the probability that, given two random species, we correctly rank their relative risks of extinction.

@@

For this exercise we chose to analyze one of the best-sampled and studied fossil records -- the Cenozoic record of skeletonized marine planktonic microorganisms F(oraminifera, Radiolaria, Diatoms, and Coccolithophores). These data are readily available through the Neptune database, an online repository of species occurrences obtained through the Deep Sea Drilling Program and the Ocean Drilling Project \citep{Lazarus1994,SpencerCervato1999}. This database provides abundant samples in space and time, a high degree of temporal resolution for the entirety of the Cenozoic, and has an internally consistent taxonomic identification strategy -- as close to ideal data for this analysis as possible.

Rarely are we able to analyze long periods of geological time at fine resolutions -- below the 5-10 My (million year) scale. Due to substantial effort and the unique biology of the system, the microfossil provides us the unique opportunity to analyze ecological and evolutionary patterns at approximately million-year time scales. Typical ``exceptional'' fossil records tend to be of individual taxonomic groups and for rarely longer than 10 million years. The Neptune database records multiple phyla-scale taxonomic groups for over 60 million years, with incredible temporal resolution supported by the various age-models of the deep-sea cores the occurrences are recorded from -- there is no equivalent fossil record. By analyzing patterns of extinction and global occurrence at fine temporal scales, we can better elucidate how well we can predict species extinction at human-relevant scales.

@@

We analyzed microfossil occurrence information which was downloaded from the Neptune Database \url{http://www.nsb-mfn-berlin.de/nannotax} \citep{Lazarus1994,SpencerCervato1999}. All occurrence information was downloaded for calcareous nannofossils, diatoms, foraminifera, and radiolarians -- these occurrences span the entire globe between 120 and 0 million years ago (Mya). This dataset of occurrences was then filtered to just those species which have their first occurrence at most 63 Mya. This choice means that our analysis avoids those taxa which survived the K/Pg boundary, those taxa which arose just after the K/Pg boundary, and means that our occurrence histories line up with the temperature time-series which was used as a predictor of extinction as discussed below.

@@

A taxon's geographic range was calculated for each of the 1 My bins in which it occurred. Geographic range was calculated as the maximum great circle distance on an ellipsoid (i.e. the Earth) between any two occurrences of that species; this distance was measured in kilometers. Geographic range was then log-plus-one transformed, and standardized by mean-centering the data and then dividing by the standard deviation of the distribution of geographic ranges. This standardization means that a regression coefficient associated with this covariate describes the change in extinction probability per change in standard deviation of geographic range, that coefficients associated with similarly standardized covariates will be directly comparable in magnitude, and that the intercept term corresponds to the expected value of the outcome at when geographic range is its average value \citep{ARM}. Change in geographic range between observations was measured from the standardized geographic range values and was not standardized separately.

Temperature data used as covariates in this analysis are based on Magnesium/Calcium isotope ratios sourced from \citet{Cramer2011}. These elemental ratios are considered more accurate estimates of past global temperature when compared to the frequently used Oxygen isotope-based estimates; this is because Mg/Ca based estimates are not effected by ice-volume and fresh-water input (e.g. meteoric water) which can alter Oxygen isotope ratios without reflecting changes to the climate itself. This property is of particular importance for this analysis as polar ice-caps develop midway through the Cenozoic. Our data source, \citet{Cramer2011}, provides temperature estimates for every 0.1 My from 0 to 63 Mya. We binned these estimates into 1 My intervals as we did with the fossil occurrences. The temperature estimate for each 1 My interval was calculated as the mean of all estimates within that interval. Temperature was then transformed and standardized the in the same manner as geographic range. Difference in temperature between between observations was measured from the standardized temperature values and was not standardized separately.

@@

The in-sample model comparisons are for determining their relative adequacy, or a model's ability to represent the data it was fit to. Comparison between the posterior predictive estimates of in-sample AUC for each of the four models demonstrates that, overall, all of the models have approximately equal in-sample performance (Fig. \ref{fig:auc_hist}). The parameter rich model VP has the greatest median in-sample AUC when compared to the other three models, but there is substantial overlap in their posterior distributions. Additionally, while our parameter rich model VP is possibly the most adequately performing model, the difference or improvement to performance is minimal at best -- all four models have approximately equal in-sample AUC posterior distributions. All of the in-sample AUC estimates from our models are concentrated around an AUC of 0.77 which is interpreted as ``fine but not good'' performance. It is therefore hard to conclude that there is one ``best'' model which we can rely upon.

@@

In survival analysis, the hazard function describes the instantaneous rate of extinction of a species given its age and relevant covariates. The hazard function is defined as the conditional probability of a species going extinct by the end of the \(t\)-th interval given that it survived up until \(t\) and the relevant covariate information \(X\) for all \(k\) 1 My intervals \citep{Tutz2016}. For the discrete time intervals \(T = 1, \cdots, k\), extinction is defined as \(T = t\). The discrete time hazard function is defined as
\begin{equation}
  \lambda(t | X) = P(T = t | T \geq t, X), \quad t = 1, \cdots, k.
  \label{eq:hazard}
\end{equation}
